Preprocessing and exploration of the narwhal data for the analysis with smoothSDE

### 2025-01-19 R/preprocessing/narwhals/data_preprocessing.Rmd
redo preprocessing with shrunk polygons land to include new covariates DistanceShore_translated and theta_translated

# Read the raw data

## Read

@@

# First data exploration

## Trials and intertrials periods

@@

#plot with all the narwhals rescaled so that the experiments start at the same time 
trials_rescaled=ggplot(data[!is.na(data$Type), ], aes(time/24,ID,fill=Type)) + 
  xlab("Number of days after tagging")+ylab("Exposure")+
  geom_tile(lwd = 1,linetype = 1)+ 
  scale_fill_manual(values=c("blue","red"),na.value="white",
                    name="Trial type",labels=c("Intertrial","Trial"))+
  scale_y_discrete(labels=c("A1", "A2", "A3", "A4","A5", "A6"))

#plot with all the narwhals at the real experiments time
trials=ggplot(data[!is.na(data$Type), ], aes(Posixct,ID,fill=Type)) + 
  xlab("Day")+ylab("Exposure")+
  geom_tile(lwd = 1.5,linetype = 1)+ 
   scale_fill_manual(values=c("blue","red"), na.value="white",
                     name="Trial type",labels=c("Intertrial","Trial"))+
  scale_y_discrete(labels=c("A1", "A2", "A3", "A4","A5", "A6"))

@@

```{r ggplot ship exposure}
#plot exposure
pExpAll=ggplot(data,aes(time,ExpShip,col=ID))+geom_line()+
  scale_color_viridis_d(labels=c("A1","A2","A3","A4","A5","A6"))

#plot exposure during intertrial
pExpInt=ggplot(data,aes(time,ExpShipI,col=ID))+geom_line()+
  scale_color_viridis_d(labels=c("A1","A2","A3","A4","A5","A6"))

#plot exposure during trial
pExpTr=ggplot(data,aes(time,ExpShipT,col=ID))+geom_line()+
  scale_color_viridis_d(labels=c("A1","A2","A3","A4","A5","A6"))

@@

```{r histogram horizontal speed,warnings=FALSE}
speed_histo=ggplot(dataprep, aes(x =horizontal_speed,col=ID)) +
  geom_histogram(fill="white",bins=50)+facet_grid(ID ~.,labeller=as_labeller(c("Asgeir"="A1","Frederik"="A2","Helge18"="A3","Kyrri"="A4","Nemo"="A5","Siggi"="A6")))+
  scale_color_viridis_d(guide="none")+
  theme(legend.position="top")+xlab("Empirical horizontal speed (km/h)")

@@

```{r histogram filtrered horizontal speed,warnings=FALSE}
speed_histo_filtered=ggplot(dataprep, aes(x =horizontal_speed,col=ID)) +
  geom_histogram(fill="white",bins=50)+facet_grid(ID ~.,labeller=as_labeller(c("Asgeir"="A1","Frederik"="A2","Helge18"="A3","Kyrri"="A4","Nemo"="A5","Siggi"="A6")))+
  scale_color_viridis_d(guide="none")+
  theme(legend.position="top")+xlab("Empirical horizontal speed (km/h)")

@@

```{r time steps histogram}
times_histo=ggplot(dtimes, aes(x =60*delta,col=ID)) +
  geom_histogram(fill="white",bins=100)+facet_grid(ID ~.,labeller=as_labeller(c("Asgeir"="A1","Frederik"="A2","Helge18"="A3","Kyrri"="A4","Nemo"="A5","Siggi"="A6")))+
  scale_color_viridis_d(guide="none")+
  theme(legend.position="top")+xlab("Time step (min) ")

@@

```{r display hist , fig.cap="Histogram of time steps",warnings=FALSE}
all_times_histo=ggplot(data=dtimes,aes(x=60*delta))+geom_histogram(bins=120,color="black",fill="grey")+xlab("Time step (min)")
all_times_histo
ggsave(path=plots_folder,"all_time_steps_histo.png", plot = all_times_histo)
```

@@

## Time steps distribution during exposure vs no exposure

@@

#static map
map=milne_land_map + 
  geom_point(dataprep[,c("ID","Long","Lat","time")],
             mapping=aes(Long, Lat, col = ID),size= 0.5,alpha=1) +
   geom_path(dataprep[,c("ID","Long","Lat","time")],mapping=aes(Long,Lat,col=ID),linewidth=0.1)+
  geom_point(data = dataprep[,c("ID","Long","Lat","time")] %>% filter(!duplicated(ID)),
             aes(x = Long, y = Lat,col=ID), shape = 3, size = 4,col = "red")+
  scale_color_viridis_d(labels=c("A1","A2","A3","A4","A5","A6"))

@@

We read the preprocessed land polygons extracted from OpenStreetMap database.

@@

```{r summary distances to shore}
summary(res_land$min_distances)
summary(dataprep$DistanceShore)
quantile(res_land$min_distance,probs=c(0.05,0.95))
D_low=quantile(res_land$min_distance,probs=0.05)
D_up=median(res_land$min_distance)
```

```{r histograms distances to shore}
histo1=ggplot(dataprep[dataprep$DistanceShore>0,], aes(x = DistanceShore)) +
  geom_histogram(fill="cyan",bins=100,color="black")+xlab("Original distance to shore (km)")

df=data.frame(DistanceShore=res_land$min_distances[res_land$min_distances>0])
histo2=ggplot(df, aes(x =DistanceShore)) +
  geom_histogram(fill="cyan",bins=100,color="black")+xlab("Computed distance fom land (km)")
```

@@

```{=tex}

For the rest of the analysis, we'll keep the bounded domain defined by the updated polygons in "updated_scoresby_sound_utm" sf object as a reference.
Doing that involves having some points on land, that is points with Distance to shore equal to 0.
```

@@

@@

@@

dataprep$AngleShore=rep(NA,n)
dataprep$AngleVelocities=rep(NA,n)
dataprep$AngleNormal=rep(NA,n)
dataprep$VelocityPhase=rep(NA,n)

@@

  #matrix of normal vectors
  normal=as.matrix(sub_data[2:n_sub,c("nx","ny")])

  #angle between velocity and normal vector
  theta_coast=signed_angle(vexp_df,normal)

  #deviation angle between consecutive velocities
  theta_vel=signed_angle(vexp_df[1:(n_sub-2),],vexp_df[2:(n_sub-1),])

@@

  #deviation from trajectory parallel to coastline
  phi=ifelse(theta_coast > 0, theta_coast - pi/2, theta_coast + pi/2)

  #adjust lengths by adding NA when necessary
  theta_vel=c(NA,theta_vel,NA)
  phi=c(NA,phi)
  theta=c(NA,theta_coast)
  vel_phase=c(NA,vel_phase,NA)

  dataprep[sub_ind,"AngleShore"]=phi
  dataprep[sub_ind,"AngleVelocities"]=theta_vel
  dataprep[sub_ind,"AngleNormal"]=-theta
  dataprep[sub_ind,"VelocityPhase"]=vel_phase
}

@@

#plot with highlight on the periods when the narwhals are exposed
plot_speed=ggplot()+ geom_line(data=dataprep,aes(time,horizontal_speed,col=ID))+ 
  geom_rect(data = regions, 
            aes(xmin = start_time, xmax = end_time, ymin = -Inf, ymax = Inf),
            alpha = 0.3,col="lightblue")+
  facet_grid(ID ~.,labeller=as_labeller(c("Asgeir"="A1","Frederik"="A2","Helge18"="A3","Kyrri"="A4","Nemo"="A5","Siggi"="A6")))+
  scale_color_viridis_d(guide="none")+theme(legend.position="none")+xlab("Time (h)")+ylab("Empirical velocity")

plot_AngleShore=ggplot()+geom_line(data=dataprep[dataprep$DistanceShore>0.05,],aes(time,AngleShore,col=ID))+
  geom_rect(data = regions, aes(xmin = start_time, xmax = end_time, ymin = -Inf, ymax = Inf),alpha = 0.3,col="lightblue")+
  facet_grid(ID ~.,labeller=as_labeller(c("Asgeir"="A1","Frederik"="A2","Helge18"="A3","Kyrri"="A4","Nemo"="A5","Siggi"="A6")))+
  scale_color_viridis_d(guide="none")+theme(legend.position="none")+xlab("Time (h)")+ylab("Empirical deviation from coastline")

plot_AngleVelocities=ggplot()+geom_line(data=dataprep,aes(time,AngleVelocities,col=ID))+
  geom_rect(data = regions, aes(xmin = start_time, xmax = end_time, ymin = -Inf, ymax = Inf),alpha = 0.3,col="lightblue")+
  facet_grid(ID ~.,labeller=as_labeller(c("Asgeir"="A1","Frederik"="A2","Helge18"="A3","Kyrri"="A4","Nemo"="A5","Siggi"="A6")))+
  scale_color_viridis_d(guide="none")+theme(legend.position="none")+xlab("Time (h)")+ylab("Empirical deviation between consecutive velocities")

plot_AngleNormal=ggplot()+geom_line(data=dataprep[dataprep$DistanceShore>0.05,],aes(time,AngleNormal,col=ID))+
  geom_rect(data = regions, aes(xmin = start_time, xmax = end_time, ymin = -Inf, ymax = Inf),alpha = 0.3,col="lightblue")+
  facet_grid(ID ~.,labeller=as_labeller(c("Asgeir"="A1","Frederik"="A2","Helge18"="A3","Kyrri"="A4","Nemo"="A5","Siggi"="A6")))+
  scale_color_viridis_d(guide="none")+theme(legend.position="none")+xlab("Time (h)")+ylab("Angle between empirical velocity and normal vector")

@@

AngleShore_histo=ggplot(dataprep[dataprep$DistanceShore>0.05,], aes(x = AngleShore,col=ID)) +
  geom_histogram(fill="white",bins=50)+facet_grid(ID ~.,labeller=as_labeller(c("Asgeir"="A1","Frederik"="A2","Helge18"="A3","Kyrri"="A4","Nemo"="A5","Siggi"="A6")))+
  scale_color_viridis_d(guide="none")+
  theme(legend.position="top")+xlab("Empirical deviation from coastline (rad)")

AngleVelocities_histo=ggplot(dataprep, aes(x = AngleVelocities,col=ID)) +
  geom_histogram(fill="white",bins=50)+facet_grid(ID ~.,labeller=as_labeller(c("Asgeir"="A1","Frederik"="A2","Helge18"="A3","Kyrri"="A4","Nemo"="A5","Siggi"="A6")))+
  scale_color_viridis_d(guide="none")+
  theme(legend.position="top")+xlab("Empirical deviation between consecutive velocities (rad) ")

AngleNormal_histo=ggplot(dataprep[dataprep$DistanceShore>0.05,], aes(x = AngleNormal,col=ID)) +
  geom_histogram(fill="white",bins=50)+facet_grid(ID ~.,labeller=as_labeller(c("Asgeir"="A1","Frederik"="A2","Helge18"="A3","Kyrri"="A4","Nemo"="A5","Siggi"="A6")))+
  scale_color_viridis_d(guide="none")+
  theme(legend.position="top")+xlab("Angle between empirical velocity and normal vector (rad) ")

VelocityPhase_histo=ggplot(dataprep, aes(x = VelocityPhase,col=ID)) +
  geom_histogram(fill="white",bins=50)+facet_grid(ID ~.,labeller=as_labeller(c("Asgeir"="A1","Frederik"="A2","Helge18"="A3","Kyrri"="A4","Nemo"="A5","Siggi"="A6")))+
  scale_color_viridis_d(guide="none")+
  theme(legend.position="top")+xlab("Velocity phase angele (rad) ")

@@

We also plot the density of the Dist_Ship variable to understand better its distribution.

@@

dataBE1=data.frame()
for (id in unique(data$ID)) {
  narwhalData=dataprep[dataprep$ID==id,]
  maxtime <- min(narwhalData$time[(narwhalData$LOS == 1)])
  dataBE1 =rbind(dataBE1,subset(narwhalData, !(time > (maxtime-1)) & (time>tagtime)))
}

dataBE2=data.frame()
for (id in unique(data$ID)) {
  narwhalData=dataprep[dataprep$ID==id,]
  maxtime <- min(narwhalData$time[(narwhalData$LOS == 1)])
  dataBE2 =rbind(dataBE2,subset(narwhalData, !(time > (maxtime-1)) & (time>tagtime/2)))
}

@@

```{r}
#number of data points for each individual
for (id in unique(data$ID)) {
  narwhalDataBE=dataBE2[dataBE2$ID==id,]
  narwhalDataAE=dataAE[dataAE$ID==id,]
  cat(id," : ", length(narwhalDataBE$time)," GPS measurement before exposure and ", length(narwhalDataAE$time)," after exposure.\n",sep="")
}
```

@@

```{r plots before and after exposure}
pBE1=milne_land_map + 
  geom_point(dataBE1[,c("ID","Long","Lat","time")],
             mapping=aes(Long, Lat, col = ID),size= 0.5,alpha=1) +
   geom_path(dataBE1[,c("ID","Long","Lat","time")],mapping=aes(Long,Lat,col=ID),linewidth=0.1)+
  geom_point(data = dataBE1[,c("ID","Long","Lat","time")] %>% filter(!duplicated(ID)),
             aes(x = Long, y = Lat,col=ID), shape = 3, size = 4,col = "red")+
  scale_color_viridis_d(labels=c("A1","A2","A3","A4","A5","A6"))

pBE2=milne_land_map + 
  geom_point(dataBE2[,c("ID","Long","Lat","time")],
             mapping=aes(Long, Lat, col = ID),size= 0.5,alpha=1) +
   geom_path(dataBE2[,c("ID","Long","Lat","time")],mapping=aes(Long,Lat,col=ID),linewidth=0.1)+
  geom_point(data = dataBE2[,c("ID","Long","Lat","time")] %>% filter(!duplicated(ID)),
             aes(x = Long, y = Lat,col=ID), shape = 3, size = 4,col = "red")+
  scale_color_viridis_d(labels=c("A1","A2","A3","A4","A5","A6"))

@@

ggsave(path=plots_folder,"tracksBE1.png",pBE1,width=10,height=5)
ggsave(path=plots_folder,"tracksBE2.png",pBE2,width=10,height=5)
ggsave(path=plots_folder,"tracksAE.png",pAE,width=10,height=5)
```

```{r display pBE1, fig.cap="Tracks before exposure, 24h after tagging"}
pBE1
```

```{r display pBE2, fig.cap="Tracks before exposure, 12h after tagging"}
pBE2
```

@@

write.csv(dataprep,file.path(preprocessed_narwhal_data_path,"allData.csv"))
write.csv(dataBE1,file.path(preprocessed_narwhal_data_path,"DataBE1.csv"))
write.csv(dataBE2,file.path(preprocessed_narwhal_data_path,"DataBE2.csv"))
write.csv(dataAE,file.path(preprocessed_narwhal_data_path,"DataAE.csv"))
```

>>>>

# Explore raw data

## Read
Takes a few minutes to read the whole dataset.

@@

# Trials and intertrials periods exploration

@@

#plot with all the narwhals rescaled so that the experiments start at the same time 
trials_rescaled=ggplot(data[!is.na(data$Type), ], aes(time/24,ID,fill=Type)) + 
  xlab("Number of days after tagging")+ylab("Exposure")+
  geom_tile(lwd = 1,linetype = 1)+ 
  scale_fill_manual(values=c("blue","red"),na.value="white",
                    name="Trial type",labels=c("Intertrial","Trial"))+
  scale_y_discrete(labels=c("A1", "A2", "A3", "A4","A5", "A6"))+theme_minimal()

#plot with all the narwhals at the real experiments time
trials=ggplot(data[!is.na(data$Type), ], aes(Posixct,ID,fill=Type)) + 
  xlab("Day")+ylab("Exposure")+
  geom_tile(lwd = 1.5,linetype = 1)+ 
   scale_fill_manual(values=c("blue","red"), na.value="white",
                     name="Trial type",labels=c("Intertrial","Trial"))+
  scale_y_discrete(labels=c("A1", "A2", "A3", "A4","A5", "A6"))+theme_minimal()

@@

```{r ggplot ship exposure}
#plot exposure
pExpAll=ggplot(data,aes(time,ExpShip,col=ID))+geom_line()+
  scale_color_viridis_d(labels=c("A1","A2","A3","A4","A5","A6"))+theme_minimal()

#plot exposure during intertrial
pExpInt=ggplot(data,aes(time,ExpShipI,col=ID))+geom_line()+
  scale_color_viridis_d(labels=c("A1","A2","A3","A4","A5","A6"))+theme_minimal()

#plot exposure during trial
pExpTr=ggplot(data,aes(time,ExpShipT,col=ID))+geom_line()+
  scale_color_viridis_d(labels=c("A1","A2","A3","A4","A5","A6"))+theme_minimal()

@@

```{r histogram horizontal speed,warnings=FALSE}
speed_histo=ggplot(dataprep, aes(x =horizontal_speed,col=ID)) +
  geom_histogram(fill="white",bins=50)+facet_grid(ID ~.,labeller=as_labeller(c("Asgeir"="A1","Frederik"="A2","Helge18"="A3","Kyrri"="A4","Nemo"="A5","Siggi"="A6")))+
  scale_color_viridis_d(guide="none")+
  theme(legend.position="top")+xlab("Empirical horizontal speed (km/h)")+theme_minimal()

@@

```{r histogram filtrered horizontal speed,warnings=FALSE}
speed_histo_filtered=ggplot(dataprep, aes(x =horizontal_speed,col=ID)) +
  geom_histogram(fill="white",bins=50)+facet_grid(ID ~.,labeller=as_labeller(c("Asgeir"="A1","Frederik"="A2","Helge18"="A3","Kyrri"="A4","Nemo"="A5","Siggi"="A6")))+
  scale_color_viridis_d(guide="none")+
  theme(legend.position="top")+xlab("Empirical horizontal speed (km/h)")+theme_minimal()

@@

```{r time steps histogram}
times_histo=ggplot(dtimes, aes(x =60*delta,col=ID)) +
  geom_histogram(fill="white",bins=100)+facet_grid(ID ~.,labeller=as_labeller(c("Asgeir"="A1","Frederik"="A2","Helge18"="A3","Kyrri"="A4","Nemo"="A5","Siggi"="A6")))+
  scale_color_viridis_d(guide="none")+
  theme(legend.position="top")+xlab("Time step (min) ")+theme_minimal()

@@

```{r display hist , fig.cap="Histogram of time steps",warnings=FALSE}
all_times_histo=ggplot(data=dtimes,aes(x=60*delta))+geom_histogram(bins=120,color="black",fill="grey")+xlab("Time step (min)")+
  theme_minimal()
all_times_histo
ggsave(path=plots_folder,"all_time_steps_histo.png", plot = all_times_histo)
```

@@

```{r}
all_times_histo_cropped=ggplot(data=dtimes[dtimes$delta<1,],aes(x=60*delta))+geom_histogram(bins=120,color="black",fill="grey")+xlab("Time step (min)")+
  theme_minimal()
all_times_histo_cropped
ggsave(path=plots_folder,"all_time_steps_histo_cropped.png", plot = all_times_histo_cropped)
```

@@

#static map
map=milne_land_map + 
  geom_point(dataprep[,c("ID","Long","Lat","time")],
             mapping=aes(Long, Lat, col = ID),size= 0.5,alpha=1) +
   geom_path(dataprep[,c("ID","Long","Lat","time")],mapping=aes(Long,Lat,col=ID),linewidth=0.1)+
  geom_point(data = dataprep[,c("ID","Long","Lat","time")] %>% filter(!duplicated(ID)),
             aes(x = Long, y = Lat,col=ID), shape = 3, size = 4,col = "red")+
  scale_color_viridis_d(labels=c("A1","A2","A3","A4","A5","A6"))+theme_minimal()

@@

We read the preprocessed land polygons extracted from OpenStreetMap database without modification except for cropping the polygons in the right bounding box and projecting it in EPSG 32626 to get UTM coordinates in km.

@@

```{r summary distances to shore}
summary(res_land$min_distances)
summary(dataprep$DistanceShore)
```

```{r histograms distances to shore}
histo1=ggplot(dataprep[dataprep$DistanceShore>0,], aes(x = DistanceShore)) +
  geom_histogram(fill="cyan",bins=100,color="black")+xlab("Original distance to shore (km)")+
  theme_minimal()

df=data.frame(DistanceShore=res_land$min_distances[res_land$min_distances>0])
histo2=ggplot(df, aes(x =DistanceShore)) +
  geom_histogram(fill="cyan",bins=100,color="black")+xlab("Computed distance fom land (km)")+
  theme_minimal()
```

@@

```{=tex}
A lot of points are indeed still very close to land. We think that this means the shoreline is still not accurate. To solve the problem, we translate the polygon vertices 200m inward. We compute new distances to shore based on the translated polygon.

```{r get shrunk polygons}

#get updated land polygons
shrunk_scoresby_sound_utm<-st_read(file.path(preprocessed_greenland_data_path,"shrunk_updated_scoresby_sound_utm.shp"))

shrunk_scoresby_sound_utm <- st_transform(shrunk_scoresby_sound_utm, crs = "+init=EPSG:32626 +units=km")

```{r compute updated translated distance to shore}
res_shrunk_land=nearest_points(positions=narwhal_positions,border=shrunk_scoresby_sound_utm)

cat("Percentage of points on land with distance from shrunk updated osm land polygons : \n",
    length(which(res_shrunk_land$min_distances==0))/length(dataprep$time)*100,"\n")

df=data.frame(DistanceShore=res_shrunk_land$min_distances[res_shrunk_land$min_distances<1 & res_shrunk_land$min_distances>0  ])
histo=ggplot(df, aes(x =DistanceShore)) +
  geom_histogram(fill="cyan",bins=100,color="black")+xlab("Computed distance to shore (km)")

@@

#same for shrunk polygons
nearest_points_shrunk=res_shrunk_land$nearest_points
min_distances_shrunk=res_shrunk_land$min_distances

#add columns px and py with coordinates of the nearest point on the shore line
dataprep$px_translated=nearest_points_shrunk[,1]
dataprep$py_translated=nearest_points_shrunk[,2]

# add columns with difference between narwhal position and nearest point on the shoreline
dataprep$nx_translated=dataprep$x-dataprep$px_translated
dataprep$ny_translated=dataprep$y-dataprep$py_translated
dataprep$DistanceShore_translated=min_distances_shrunk
```

@@

#add Ishore covariate based on 1/distance in km
dataprep$Ishore_translated <- ExposureFunction(X = dataprep$DistanceShore_translated)

@@

dataprep$AngleShore=rep(NA,n)
dataprep$AngleVelocities=rep(NA,n)
dataprep$theta=rep(NA,n)
dataprep$theta_translated=rep(NA,n)
dataprep$VelocityPhase=rep(NA,n)

@@

  #matrix of normal vectors
  normal=as.matrix(sub_data[2:n_sub,c("nx","ny")])
  normal_translated=as.matrix(sub_data[2:n_sub,c("nx_translated","ny_translated")])

  #angle between velocity and normal vector
  theta=signed_angle(vexp_df,normal)
  theta_translated=signed_angle(vexp_df,normal_translated)

  #deviation angle between consecutive velocities
  angle_vel=signed_angle(vexp_df[1:(n_sub-2),],vexp_df[2:(n_sub-1),])

@@

  #deviation from trajectory parallel to coastline
  phi=ifelse(theta > 0, theta - pi/2, theta + pi/2)

  #adjust lengths by adding NA when necessary
  angle_vel=c(NA,angle_vel,NA)
  phi=c(NA,phi)
  theta=c(NA,theta)
  theta_translated=c(NA,theta_translated)
  vel_phase=c(NA,vel_phase,NA)

  dataprep[sub_ind,"AngleShore"]=phi
  dataprep[sub_ind,"AngleVelocities"]=angle_vel
  dataprep[sub_ind,"theta"]=-theta
  dataprep[sub_ind,"theta_translated"]=-theta_translated
  dataprep[sub_ind,"VelocityPhase"]=vel_phase
}

@@

#plot with highlight on the periods when the narwhals are exposed
plot_speed=ggplot()+ geom_line(data=dataprep,aes(time,horizontal_speed,col=ID))+ 
  geom_rect(data = regions, 
            aes(xmin = start_time, xmax = end_time, ymin = -Inf, ymax = Inf),
            alpha = 0.3,col="lightblue")+
  facet_grid(ID ~.,labeller=as_labeller(c("Asgeir"="A1","Frederik"="A2","Helge18"="A3","Kyrri"="A4","Nemo"="A5","Siggi"="A6")))+
  scale_color_viridis_d(guide="none")+theme(legend.position="none")+xlab("Time (h)")+ylab("Empirical velocity")+
   theme_minimal()

plot_AngleShore=ggplot()+geom_line(data=dataprep[dataprep$DistanceShore>0.05,],aes(time,AngleShore,col=ID))+
  geom_rect(data = regions, aes(xmin = start_time, xmax = end_time, ymin = -Inf, ymax = Inf),alpha = 0.3,col="lightblue")+
  facet_grid(ID ~.,labeller=as_labeller(c("Asgeir"="A1","Frederik"="A2","Helge18"="A3","Kyrri"="A4","Nemo"="A5","Siggi"="A6")))+
  scale_color_viridis_d(guide="none")+theme(legend.position="none")+xlab("Time (h)")+ylab("Empirical deviation from coastline")+
  theme_minimal()

plot_AngleVelocities=ggplot()+geom_line(data=dataprep,aes(time,AngleVelocities,col=ID))+
  geom_rect(data = regions, aes(xmin = start_time, xmax = end_time, ymin = -Inf, ymax = Inf),alpha = 0.3,col="lightblue")+
  facet_grid(ID ~.,labeller=as_labeller(c("Asgeir"="A1","Frederik"="A2","Helge18"="A3","Kyrri"="A4","Nemo"="A5","Siggi"="A6")))+
  scale_color_viridis_d(guide="none")+theme(legend.position="none")+xlab("Time (h)")+ylab("Empirical deviation between consecutive velocities")+ theme_minimal()

plot_AngleNormal=ggplot()+geom_line(data=dataprep[dataprep$DistanceShore>0.05,],aes(time,theta,col=ID))+
  geom_rect(data = regions, aes(xmin = start_time, xmax = end_time, ymin = -Inf, ymax = Inf),alpha = 0.3,col="lightblue")+
  facet_grid(ID ~.,labeller=as_labeller(c("Asgeir"="A1","Frederik"="A2","Helge18"="A3","Kyrri"="A4","Nemo"="A5","Siggi"="A6")))+
  scale_color_viridis_d(guide="none")+theme(legend.position="none")+xlab("Time (h)")+ylab("Angle between empirical velocity and normal vector")+ theme_minimal()

@@

AngleShore_histo=ggplot(dataprep[dataprep$DistanceShore>0.05,], aes(x = AngleShore,col=ID)) +
  geom_histogram(fill="white",bins=50)+facet_grid(ID ~.,labeller=as_labeller(c("Asgeir"="A1","Frederik"="A2","Helge18"="A3","Kyrri"="A4","Nemo"="A5","Siggi"="A6")))+
  scale_color_viridis_d(guide="none")+
  theme(legend.position="top")+xlab("Empirical deviation from coastline (rad)")+ theme_minimal()

AngleVelocities_histo=ggplot(dataprep, aes(x = AngleVelocities,col=ID)) +
  geom_histogram(fill="white",bins=50)+facet_grid(ID ~.,labeller=as_labeller(c("Asgeir"="A1","Frederik"="A2","Helge18"="A3","Kyrri"="A4","Nemo"="A5","Siggi"="A6")))+
  scale_color_viridis_d(guide="none")+
  theme(legend.position="top")+xlab("Empirical deviation between consecutive velocities (rad) ")+ theme_minimal()

AngleNormal_histo=ggplot(dataprep[dataprep$DistanceShore>0.05,], aes(x = theta,col=ID)) +
  geom_histogram(fill="white",bins=50)+facet_grid(ID ~.,labeller=as_labeller(c("Asgeir"="A1","Frederik"="A2","Helge18"="A3","Kyrri"="A4","Nemo"="A5","Siggi"="A6")))+
  scale_color_viridis_d(guide="none")+
  theme(legend.position="top")+xlab("Angle between empirical velocity and normal vector (rad) ")+ theme_minimal()

VelocityPhase_histo=ggplot(dataprep, aes(x = VelocityPhase,col=ID)) +
  geom_histogram(fill="white",bins=50)+facet_grid(ID ~.,labeller=as_labeller(c("Asgeir"="A1","Frederik"="A2","Helge18"="A3","Kyrri"="A4","Nemo"="A5","Siggi"="A6")))+
  scale_color_viridis_d(guide="none")+
  theme(legend.position="top")+xlab("Velocity phase angele (rad) ")+ theme_minimal()

@@

We also plot the histogram of the Dist_Ship variable to understand better its distribution.

@@

dataBE24=data.frame()
for (id in unique(data$ID)) {
  narwhalData=dataprep[dataprep$ID==id,]
  maxtime <- min(narwhalData$time[(narwhalData$LOS == 1)])
  dataBE24 =rbind(dataBE24,subset(narwhalData, !(time > (maxtime-1)) & (time>tagtime)))
}

dataBE12=data.frame()
for (id in unique(data$ID)) {
  narwhalData=dataprep[dataprep$ID==id,]
  maxtime <- min(narwhalData$time[(narwhalData$LOS == 1)])
  dataBE12 =rbind(dataBE12,subset(narwhalData, !(time > (maxtime-1)) & (time>tagtime/2)))
}

@@

```{r}
#number of data points for each individual
for (id in unique(data$ID)) {
  narwhalDataBE=dataBE12[dataBE12$ID==id,]
  narwhalDataAE=dataAE[dataAE$ID==id,]
  cat(id," : ", length(narwhalDataBE$time)," GPS measurement before exposure and ", length(narwhalDataAE$time)," after exposure.\n",sep="")
}
```

@@

```{r plots before and after exposure}
pBE24=milne_land_map + 
  geom_point(dataBE24[,c("ID","Long","Lat","time")],
             mapping=aes(Long, Lat, col = ID),size= 0.5,alpha=1) +
   geom_path(dataBE24[,c("ID","Long","Lat","time")],mapping=aes(Long,Lat,col=ID),linewidth=0.1)+
  geom_point(data = dataBE24[,c("ID","Long","Lat","time")] %>% filter(!duplicated(ID)),
             aes(x = Long, y = Lat,col=ID), shape = 3, size = 4,col = "red")+
  scale_color_viridis_d(labels=c("A1","A2","A3","A4","A5","A6"))

pBE12=milne_land_map + 
  geom_point(dataBE12[,c("ID","Long","Lat","time")],
             mapping=aes(Long, Lat, col = ID),size= 0.5,alpha=1) +
   geom_path(dataBE12[,c("ID","Long","Lat","time")],mapping=aes(Long,Lat,col=ID),linewidth=0.1)+
  geom_point(data = dataBE12[,c("ID","Long","Lat","time")] %>% filter(!duplicated(ID)),
             aes(x = Long, y = Lat,col=ID), shape = 3, size = 4,col = "red")+
  scale_color_viridis_d(labels=c("A1","A2","A3","A4","A5","A6"))

@@

ggsave(path=plots_folder,"tracksBE12.png",pBE12,width=10,height=5)
ggsave(path=plots_folder,"tracksBE24.png",pBE24,width=10,height=5)
ggsave(path=plots_folder,"tracksAE.png",pAE,width=10,height=5)
```

```{r display pBE1, fig.cap="Tracks before exposure, 24h after tagging"}
pBE12
```

```{r display pBE2, fig.cap="Tracks before exposure, 12h after tagging"}
pBE24
```

@@

write.csv(dataprep,file.path(preprocessed_narwhal_data_path,"allData.csv"))
write.csv(dataBE12,file.path(preprocessed_narwhal_data_path,"DataBE12.csv"))
write.csv(dataBE24,file.path(preprocessed_narwhal_data_path,"DataBE24.csv"))
write.csv(dataAE,file.path(preprocessed_narwhal_data_path,"DataAE.csv"))
```
